---
title: "ISyE 6414 Final Project: Predicting for Near-Earth Comets"
author: "Troy Allen, Roy Gabriel, Nattakorn Kittisut, Corey Smith & Moe Kyaw Thu"
date: "2024-06-13"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 6
)
```

# Setup

## Load Libraries

```{r load-libraries}
# Load MASS before tidyverse so dplyr::select is not masked
library(MASS)
library(tidyverse)
library(caret)
library(glmnet)
library(stargazer)
library(car)
library(lmtest)
library(gghalfnorm)
library(faraway)
```

## Load Data

```{r load-data}
comet_data <- read.csv("Near-Earth_Comets_-_Orbital_Elements_20240605.csv")
head(comet_data)
```

# Data Cleaning

## Missing Values

```{r missing-values}
missing_values <- comet_data %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "Column", values_to = "MissingValues") %>%
  filter(MissingValues > 0)

missing_values
```

The columns with missing values are:

- **A1**: 122 missing values
- **A2**: 119 missing values
- **A3**: 150 missing values
- **DT**: 156 missing values

Given the high proportion of missing values, dropping these columns is preferable to
dropping rows.

```{r inspect-missing-columns}
comet_data[, c("A1..AU.d.2.", "A2..AU.d.2.", "A3..AU.d.2.", "DT..d.")] %>%
  head(10)
```

## Drop Columns with Excessive Missing Data

We drop A1, A2, A3, and DT due to the high number of missing values.

```{r drop-columns}
comet_data_clean <- comet_data %>%
  select(-c("A1..AU.d.2.", "A2..AU.d.2.", "A3..AU.d.2.", "DT..d."))

head(comet_data_clean)
```

## Outlier Detection

Using Z-scores with a threshold of 2 to identify outliers across numeric columns.

```{r outlier-detection}
numeric_cols <- select_if(comet_data_clean, is.numeric)
z_scores <- numeric_cols %>% mutate(across(everything(), scale))

threshold <- 2

outliers <- comet_data_clean %>%
  filter(rowSums(abs(z_scores) > threshold, na.rm = TRUE) > 0)

cat("Number of outlier rows:", nrow(outliers), "\n")
```

```{r outlier-columns}
outlier_columns <- apply(abs(z_scores) > threshold, 1, function(x) colnames(comet_data_clean)[x])

for (i in 1:nrow(outliers)) {
  if (length(outlier_columns[[i]]) > 0) {
    cat("Row:", i, "\n")
    cat("Outlier columns:", paste(outlier_columns[i], collapse = ", "), "\n")
    print(outliers[i, ])
    cat("\n")
  }
}
```

## Summary Statistics

```{r summary-stats}
numeric_data <- comet_data_clean[, sapply(comet_data_clean, is.numeric)]

summary_stats <- data.frame(
  Mean = colMeans(numeric_data, na.rm = TRUE),
  SD = sapply(numeric_data, sd, na.rm = TRUE)
)

round(summary_stats, 4)
```

## Standardization

Creating a standardized version of the numeric data for potential use in modeling.

```{r standardize}
standardized_comet_data <- scale(numeric_data)
head(standardized_comet_data)
```

# Exploratory Data Analysis

## Eccentricity vs. Inclination

```{r eda-eccentricity-inclination}
ggplot(comet_data_clean, aes(x = e, y = i..deg.)) +
  geom_point(alpha = 0.6, color = "steelblue") +
  labs(
    x = "Eccentricity",
    y = "Inclination (deg)",
    title = "Eccentricity vs. Inclination"
  ) +
  theme_minimal()
```

## Eccentricity vs. Perihelion Distance

```{r eda-eccentricity-perihelion}
ggplot(comet_data_clean, aes(x = e, y = q..AU.)) +
  geom_point(alpha = 0.6, color = "steelblue") +
  labs(
    x = "Eccentricity",
    y = "Perihelion Distance (q, AU)",
    title = "Eccentricity vs. Perihelion Distance"
  ) +
  theme_minimal()
```

## Scatterplot Matrix of Orbital Elements

```{r eda-scatterplot-matrix, fig.width=10, fig.height=10}
orbital_elements <- comet_data[, c("q..AU.", "Epoch..TDB.", "TP..TDB.", "e",
                                    "i..deg.", "w..deg.", "Node..deg.")]

pairs(orbital_elements, pch = 21, bg = "skyblue",
      main = "Scatterplot Matrix of Orbital Elements")
```

# Research Question 1: Which Orbital Elements Best Predict MOID?

> **RQ1:** Which orbital elements best predict the minimum distance between the comet
> and Earth (MOID)? Are all input parameters significant? Is MOID impacted more by
> orbit eccentricity (e) or its critical angles (i, w, Node)?

## Full Linear Model

```{r rq1-load-cleaned-data}
orbit <- read.csv("comet_data_cleaned.csv")
sub.orbit <- orbit[, 2:11]
```

```{r rq1-full-model}
rq1.lm <- lm(MOID..AU. ~ ., data = sub.orbit)
summary(rq1.lm)
```

**Interpretation:** The significant predictors are `q.AU`, `Node.deg`, `w.deg`, and `e`.

## Model Diagnostics (Full Model)

```{r rq1-diagnostics}
# Normality
shapiro.test(resid(rq1.lm))
```

There is strong evidence that the residuals are **not normally distributed**.

```{r rq1-vif}
# Multicollinearity
vif(rq1.lm)
```

High VIF values for `Epoch.TDB`, `TP.TDB`, `Q.AU`, and `P.yr` indicate **multicollinearity**.

```{r rq1-heteroskedasticity}
# Heteroskedasticity
ncvTest(rq1.lm)
```

With p-value < 0.05, there is **heteroskedasticity** in the model.

```{r rq1-cooks-distance}
# Influential observations
cook.rq.1 <- cooks.distance(rq1.lm)
halfnorm(cook.rq.1, 5, ylab = "Cook's Distances")
```

Outliers identified at observations: **5, 158, 6, 22, and 160**.

## Diagnostic Summary Table

```{r rq1-diagnostic-table}
shapiro_w <- 0.89569
shapiro_p <- 3.234e-09

vif_values <- c(
  Epoch_TDB = 575.73, TP_TDB = 575.16, e = 11.27, i_deg = 2.15,
  w_deg = 1.14, Node_deg = 1.15, q_AU = 5.18, Q_AU = 152.79, P_yr = 108.35
)

ncv_chisquare <- 21.55
ncv_p <- 3.45e-06

result_table <- data.frame(
  Test = c(
    "Shapiro-Wilk Normality",
    rep("Variance Inflation Factor", length(vif_values)),
    "Non-constant Variance",
    "Cook's Distance Outliers"
  ),
  Result = c(
    paste("W:", shapiro_w, ", p-value:", format(shapiro_p, scientific = TRUE)),
    paste(names(vif_values), ":", round(vif_values, 2)),
    paste("Chi-sq:", ncv_chisquare, ", p:", format(ncv_p, scientific = TRUE)),
    "Observations: 5, 158, 6, 22, 160"
  )
)

stargazer(result_table, type = "text", summary = FALSE, rownames = FALSE)
```

## Forward Stepwise Regression

```{r rq1-forward-stepwise}
simple_model <- lm(MOID..AU. ~ 1, data = sub.orbit)
full_model <- lm(MOID..AU. ~ ., data = sub.orbit)

rq1_forward <- step(simple_model,
  scope = list(lower = simple_model, upper = full_model),
  direction = "forward", trace = 0
)

summary(rq1_forward)
```

Selected variables: `Q.AU`, `P.yr`, `q.AU`, `e`, `Node.deg`, and `w.deg` (lowest AIC: -744.09).

```{r rq1-forward-diagnostics, fig.height=8}
par(mfrow = c(2, 2))
qqnorm(residuals(rq1_forward)); qqline(residuals(rq1_forward))
hist(residuals(rq1_forward), main = "Histogram of Residuals", xlab = "Residuals")
plot(residuals(rq1_forward), xlab = "Order", ylab = "Residuals")
abline(0, 0, lty = 1, col = "red")
plot(fitted(rq1_forward), residuals(rq1_forward), xlab = "Fitted Values", ylab = "Residuals")
abline(0, 0, lty = 1, col = "red")
```

## Backward Stepwise Regression

```{r rq1-backward-stepwise}
rq1_backward <- step(full_model,
  scope = list(lower = simple_model, upper = full_model),
  direction = "backward", trace = 0
)

summary(rq1_backward)
```

Selected variables: `e`, `w.deg`, `Node.deg`, `q.AU`, and `P.yr` (lowest AIC: -746.01).

```{r rq1-backward-diagnostics, fig.height=8}
par(mfrow = c(2, 2))
qqnorm(residuals(rq1_backward)); qqline(residuals(rq1_backward))
hist(residuals(rq1_backward), main = "Histogram of Residuals", xlab = "Residuals")
plot(residuals(rq1_backward), xlab = "Order", ylab = "Residuals")
abline(0, 0, lty = 1, col = "red")
plot(fitted(rq1_backward), residuals(rq1_backward), xlab = "Fitted Values", ylab = "Residuals")
abline(0, 0, lty = 1, col = "red")
```

## Lasso Regression

```{r rq1-lasso}
X <- model.matrix(MOID..AU. ~ ., data = sub.orbit)[, -1]
y <- sub.orbit$MOID..AU.

cv_model <- cv.glmnet(X, y, alpha = 1)
best_lambda <- cv_model$lambda.min
cat("Best lambda:", best_lambda, "\n")

best_lasso <- glmnet(X, y, alpha = 1, lambda = best_lambda)
coefficients <- as.matrix(coef(best_lasso))
print(coefficients)
```

```{r rq1-lasso-plot}
base_lasso <- glmnet(X, y, alpha = 1)
plot(base_lasso, xvar = "lambda", label = TRUE)
title("Lasso Coefficient Paths")
```

```{r rq1-lasso-selection}
threshold <- 1e-6
selected_vars <- rownames(coefficients)[abs(coefficients) > threshold]
selected_vars <- selected_vars[selected_vars != "(Intercept)"]

cat("Selected variables by Lasso:\n")
print(selected_vars)
```

At the optimal lambda of ~0.00175, the selected variables are: `e`, `i.deg`, `w.deg`,
`Node.deg`, `q.AU`, and `P.yr`.

## Lasso-Selected Model

```{r rq1-lasso-model}
rq1_lasso_model <- lm(MOID..AU. ~ e + i..deg. + w..deg. + Node..deg. + q..AU. + P..yr.,
  data = sub.orbit
)

summary(rq1_lasso_model)
```

All variables except `i.deg` are significant predictors for MOID.

```{r rq1-lasso-model-diagnostics, fig.height=8}
par(mfrow = c(2, 2))
qqnorm(residuals(rq1_lasso_model)); qqline(residuals(rq1_lasso_model))
hist(residuals(rq1_lasso_model), main = "Histogram of Residuals", xlab = "Residuals")
plot(residuals(rq1_lasso_model), xlab = "Order", ylab = "Residuals")
abline(0, 0, lty = 1, col = "red")
plot(fitted(rq1_lasso_model), residuals(rq1_lasso_model), xlab = "Fitted Values", ylab = "Residuals")
abline(0, 0, lty = 1, col = "red")
```

# Research Question 2: Impact of Perihelion & Aphelion on Orbital Period

> **RQ2:** Do the minimum (q) and maximum (Q) distances between the comet and the Sun
> have an impact on the orbital period (P)? Are these values correlated or dependent on
> other orbital elements?

## Initial Model with Interaction Term

```{r rq2-initial-model}
rq2.lm <- lm(P..yr. ~ q..AU. + Q..AU. + I(q..AU. * Q..AU.), data = sub.orbit)
summary(rq2.lm)
```

The maximum distance (Q) and the q*Q interaction significantly impact orbital period,
while minimum distance (q) alone does not.

## Correlation Analysis

```{r rq2-correlation}
cat("Correlation between q and Q:", cor(sub.orbit$q..AU., sub.orbit$Q..AU.), "\n")
```

Weak positive correlation (r = 0.044) between perihelion and aphelion distances.

```{r rq2-correlation-matrix}
round(cor(sub.orbit), 2)
```

Eccentricity (`e`) has the highest correlation with both `q` and `Q`.

## Extended Model with Eccentricity

```{r rq2-extended-model}
rq2.lm.2 <- lm(P..yr. ~ q..AU. + Q..AU. + I(q..AU. * Q..AU.) + e, data = sub.orbit)
summary(rq2.lm.2)
```

With eccentricity included, **all variables become significant**.

## Model Diagnostics (RQ2)

```{r rq2-normality}
# Normality
cat("Model 1:\n"); shapiro.test(resid(rq2.lm))
cat("\nModel 2:\n"); shapiro.test(resid(rq2.lm.2))
```

Both models' residuals are **not normally distributed**.

```{r rq2-vif}
# Multicollinearity
cat("Model 1 VIF:\n"); vif(rq2.lm)
cat("\nModel 2 VIF:\n"); vif(rq2.lm.2)
```

In both models, Q and the q*Q interaction term have high VIFs.

```{r rq2-heteroskedasticity}
# Heteroskedasticity
cat("Model 1:\n"); ncvTest(rq2.lm)
cat("\nModel 2:\n"); ncvTest(rq2.lm.2)
```

Both models exhibit **heteroskedasticity** (p < 0.05).

```{r rq2-cooks-distance}
# Influential observations
par(mfrow = c(1, 2))
halfnorm(cooks.distance(rq2.lm), 5, ylab = "Cook's Distances", main = "Model 1")
halfnorm(cooks.distance(rq2.lm.2), 5, ylab = "Cook's Distances", main = "Model 2")
```

Key outliers: observations **159** and **123**.

# Research Question 3: Relationship Between MOID and Orbital Period

> **RQ3:** Is there any relationship between the two response variables, MOID and P?
> Is there a way to predict the likelihood and/or frequency of a near-Earth comet sighting?

## Simple Linear Model: MOID ~ P

```{r rq3-simple-model}
rq3.lm <- lm(MOID..AU. ~ P..yr., data = sub.orbit)
summary(rq3.lm)
```

There is a **highly significant** relationship between MOID and P, though the coefficient
is small (0.0012).

```{r rq3-correlation}
cat("Correlation between MOID and P:", cor(sub.orbit$MOID..AU., sub.orbit$P..yr.), "\n")
```

Moderate positive correlation (r ≈ 0.34).

## Extended Model: MOID ~ P + Q + e

From the RQ2 correlation analysis, MOID is positively correlated with P (0.34) and Q (0.29),
while P is strongly correlated with Q (0.98) and e (0.60).

```{r rq3-extended-model}
rq3.lm.2 <- lm(MOID..AU. ~ P..yr. + Q..AU. + e, data = sub.orbit)
summary(rq3.lm.2)
```

## Logistic Regression: Predicting Near-Earth Classification

A comet is classified as "near-Earth" if MOID < 0.05 AU.

```{r rq3-logistic-model}
sub.orbit$Near <- ifelse(sub.orbit$MOID..AU. < 0.05, 1, 0)

cat("Near-Earth comets:", sum(sub.orbit$Near), "\n")
cat("Non-near-Earth comets:", sum(sub.orbit$Near == 0), "\n")

rq3.logistic <- glm(
  Near ~ P..yr. + q..AU. + Q..AU. + e + i..deg. + w..deg. + Node..deg.,
  family = binomial,
  data = sub.orbit
)

summary(rq3.logistic)
```

Only **q**, **e**, and **Node.deg** are significant predictors of near-Earth classification.

## Model Diagnostics (RQ3)

```{r rq3-normality}
# Normality
cat("Simple model:\n"); shapiro.test(resid(rq3.lm))
cat("\nExtended model:\n"); shapiro.test(resid(rq3.lm.2))
cat("\nLogistic model:\n"); shapiro.test(resid(rq3.logistic))
```

None of the models' residuals are normally distributed.

```{r rq3-vif}
# Multicollinearity
cat("Extended model VIF:\n"); vif(rq3.lm.2)
cat("\nLogistic model VIF:\n"); vif(rq3.logistic)
```

P and Q have high VIF values in both models, indicating significant **multicollinearity**.

```{r rq3-heteroskedasticity}
# Heteroskedasticity
cat("Simple model:\n"); ncvTest(rq3.lm)
cat("\nExtended model:\n"); ncvTest(rq3.lm.2)
cat("\nLogistic model (Breusch-Pagan):\n"); bptest(rq3.logistic)
```

All three models exhibit **heteroskedasticity**.

```{r rq3-cooks-distance, fig.height=4}
# Influential observations
par(mfrow = c(1, 3))
halfnorm(cooks.distance(rq3.lm), 5, ylab = "Cook's Distances", main = "Simple Model")
halfnorm(cooks.distance(rq3.lm.2), 5, ylab = "Cook's Distances", main = "Extended Model")
halfnorm(cooks.distance(rq3.logistic), 5, ylab = "Cook's Distances", main = "Logistic Model")
```

Consistent outliers across models: observations **123** and **98**.
